---
title: "Election Prediction"
author: "Graham Tierney"
date: "10/11/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(R2jags)

```

## Set up data

Here we load the polling data and drop duplicate questions. 

```{r load,eval=FALSE}

polls <- read_csv("data/2016_Economist_Polls.csv") %>% 
  filter(population %in% c("Likely Voters","Registered Voters"),question.iteration == 1) %>% #remove duplicate questions
  mutate(days_to_election = as.Date("2016-11-08") - end.date,
         state = ifelse(state == "--","US",state),
         y = clinton/(clinton + trump)*100)

### EXERCISE: filter polls to only polls 100 days or fewer before the election and the following states: US,WI,OH,PA,NC,FL,VA
polls <- polls %>% 
  filter()
```

## Create data list

JAGS and Stan neeed the data passed in as a named list. Create that here. Some have been filled in for you

```{r data_list,eval=FALSE}
states <- polls$state %>% unique

y <- polls$y
r <- match(polls$state,states)
t <- polls$days_to_election + 1 #WHY PLUS ONE?

N_polls <- 
N_states <- 
N_days <- 

jags_data <- list(y=y,t=t,r=r,
                  N_polls=N_polls,N_states=N_states,N_days=N_days)
```

## Write model

The model we will use is $Y_k$, Clinton's share of the two-party vote, is normally distributed with mean $\beta_{jt}$, where the poll ended $t$ days before the election and was conducted on population $j$. Each state has its own observation variance, $\sigma^2_{yj}$. Thus, $Y_k \sim N(\beta_{jt} + \delta_t,\sigma^2_{yh})$. We use a random walk for $\beta$, $\beta_{jt} \sim N(\beta_{jt-1},\sigma^2_{\beta j})$. We will let other states influence each other by placing a hierarhcial prior on the model parameters: $\{\sigma^2_{yj},\sigma^2_{\beta j},\beta_{j1}\}$. In the code below, everything is set up EXCEPT these heirarchical structures. You can choosee whatever distributions you like. 

```{r model,eval = F}
model <- function(){
  for(k in 1:N_polls){
    y[k] ~ dnorm(p[k],1/sigma2_y[r[k]]) #note no longer binomial
    p[k] = beta[r[k],t[k]] 
  }
  for(j in 2:N_days){
    for(i in 1:N_states){
      beta[i,j] ~ dnorm(beta[i,j-1],pow(sigma2_beta[i],-1))
    }
  }
  
  #EXERCISE: add hierarhciacl prior for sigma2_beta and sigma2_y, i.e. sigma2_beta[j] all come from a common distribution 
  for(j in 1:N_states){
      sigma2_y[j] 
      
      sigma2_beta[j] 

      beta[j,1] 
  }

}
```

## Run and analyze results

```{r run_model,eval=F}

#be sure to add your added parameters to parameters.to.save
jags_sims <- jags(data = jags_data,model.file = model,parameters.to.save = c("beta","sigma2_beta",
                                                                             "p","sigma2_y"),
                  n.iter = 10000)


```

Now, lets look at some results. 

1) For one or two states, plot the expected value of a poll of that state ($\beta_{jt} + \delta_t$), 95\% CIs, and observed poll results. Does your model look good? 

2) Compare the implied binomial variance: $p(1-p)/n$ to the estimated variance $\sigma^2_{yj}$ for one or two states. Do the quantities look similar? 

3) Who do you predict is the winner in each state? Give a forecast electoral college vote split by awarding all non-modeled results to the actual winner. EC votes: US-0, FL-29, NC-15, OH-18, PA-20, VA-13, WI-10. 

4) How does the probability of winning one state affect another? Estimate P(Clinton wins NC) and P(Clinton wins NC|Clinton wins OH), or pick your own pair of states. Are the probabilities different? Should they be? 

5) Finally, try removing all data from the final month of the campaign from `jags_data`, then re-estimate the model. Do your predictions change? 

```{r solutions}

```

## Bonus: Multivariate

Now, instead of having all states equally correlated with each other, we will estimate the result. Now $Y_k \sim N(\theta_{jt},\sigma^2_{yj})$. The evolution is as follows: 

\[\theta_{\cdot t} \sim MVN(\theta_{\cdot t-1},\Sigma)\]

Using same priors on $\theta_{\cdot 1}$, the day of the election, as before, rewrite the jags code to implement the above model. 

```{r mv_model}
model_mv <- function(){
  for(k in 1:N_polls){
    y[k] ~ dnorm(p[k],1/sigma2_y[r[k]]) #note no longer binomial
    p[k] = theta[r[k],t[k]]
  }
  for(j in 2:N_days){
    theta[1:N_states,j] ~ dmnorm(XXX,Phi)
  }
  Phi ~ dwish(XXX,XXX) #fill in wishart parameters, google JAGS wishart distribution should turn it up
  Sigma = inverse(Phi)
  #which, Phi or Sigma, is the covariance and which is the precision? 
  
  #Use your hierarhciacl prior for sigma2_y and any theta1 prior you like
  for(j in 1:N_states){
      sigma2_y[j] 
  }
}
```

Re-estimate the model. 

```{r run_model,eval=F}

#be sure to add your added parameters to parameters.to.save
jags_sims <- jags(data = jags_data,model.file = model_mv,parameters.to.save = c("theta","Sigma",
                                                                                "p","sigma2_y"),
                  n.iter = 10000)
```

Now, how does the probability of winning one state affect another? Estimate P(Clinton wins NC) and P(Clinton wins NC|Clinton wins OH), or pick your own pair of states. Are the probabilities different? 

```{r mv_solutions}

```

Try to visualize the correlation matrix. The `get_heatmap` function below will take a correlation matrix and make a heatmap. 

```{r heatmap}

reorder_cormat <- function(cormat){
  # Use correlation between variables as distance
  dd <- as.dist((1-cormat)/2)
  hc <- hclust(dd)
  cormat <-cormat[hc$order, hc$order]
}
get_upper_tri <- function(cormat){
  cormat[lower.tri(cormat)]<- NA
  return(cormat)
}

get_heatmap <- function(cormat){
  cormat %>% 
    cov2cor() %>% 
    reorder_cormat %>% 
    get_upper_tri %>% 
    reshape2::melt(na.rm = TRUE) %>% 
    ggplot(aes(Var2, Var1, fill = value))+
    geom_tile(color = "white")+
    scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                         midpoint = 0, limit = c(-1,1), space = "Lab", 
                         name="Pearson\nCorrelation") +
    theme_minimal()+ # minimal theme
    labs(x="",y="") + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                     size = 12, hjust = 1))+
    coord_fixed()
}
```


```{r plot_cor}

```
